PAGAI

### 2012-05-05 papers/TAPAS2012/pagai.tex
TAPAS

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

%\newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
%\newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
%\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
%\newcommand{\MM}[1]{}
%\newcommand{\JH}[1]{}
%\newcommand{\DM}[1]{}

\usepackage{ifxetex}
\ifxetex
\else
\usepackage[utf8]{inputenc}
\fi

\usepackage{amsmath}
\usepackage{mathcomp,amsfonts,amssymb,MnSymbol}
\usepackage{dmnatbib}
\usepackage[mathscr]{euscript}
\usepackage{algorithmic,algorithm}
%\usepackage{graphicx}
%\usepackage{color}

@@

\usepackage{listings}
\lstset{language=C,mathescape=true,numbers=left,numbersep=0.8ex,tabsize=2,breaklines=true,basicstyle={\rm\fontfamily{phv}\small\selectfont}}

\usepackage{tikz}
\usetikzlibrary{arrows,automata}

\tikzstyle{state}=[circle,fill=black!25,minimum size=13pt,inner sep=0pt]
\tikzstyle{rstate}=[rectangle,fill=black!25,minimum size=13pt,inner sep=0pt]
\tikzstyle{transition}=[rectangle,semithick,draw=black!75,
  			  minimum size=4mm]
\tikzstyle{transition2}=[transition,rectangle,thick,dashed,
  			  minimum size=4mm]
\tikzstyle{PRstate}=[circle,double,draw,fill=blue!15,minimum size=13pt,inner sep=0pt]
\tikzstyle{polyhedra}=[blue!25,opacity=0.5,pattern=north west lines,pattern
color=blue]
\tikzstyle{line}=[black,thick]

\usepackage[pdfusetitle]{hyperref}

@@

% commentaires dans les algorithmes crees avec le package algorithmic
\renewcommand{\algorithmiccomment}[1]{ {\color{gray} // #1}}

\begin{document}

\conferenceinfo{PLDI '12}{11-16 June 2012, Beijing} 
\copyrightyear{2012} 
\copyrightdata{[to be supplied]}

%\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{Succinct representations for abstract interpretation}   % 'preprint' option specified.

\title{Succinct Representations for Abstract Interpretation}
\subtitle{Combined analysis algorithms and experimental evaluation}

\authorinfo{Anonymous}
           {The Internet}
           {anonymous.gal@gmail.com}

\maketitle

\begin{abstract}
Abstract interpretation techniques can be made more precise by distinguishing
paths inside loops, at the expense of possibly exponential complexity. SMT-solving techniques and sparse representations of paths and sets of paths avoid this pitfall.

We improve previously proposed techniques for guided static analysis and the generation of disjunctive invariants by combining them with techniques for succinct representations of paths and symbolic representations for transitions based on static single assignment.

Because of the non-monotonicity of the results of abstract interpretation with widening operators, it is difficult to conclude that some abstraction is more precise than another based on theoretical local precision results. We thus conducted extensive comparisons between our new techniques and previous ones, on a variety of open-source packages. We also took the opportunity to compare the precision of various classical abstract domains (convex polyhedra, octagons, intervals).
\end{abstract}

% \category{CR-number}{subcategory}{third-level}
\category{D.2.4}{Software/Program Verification}{Formal methods}
\category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{Assertions, Invariants, Mechanical verification, Pre- and post-conditions}
\category{F.3.2}{Semantics of Programming Languages}{Program analysis}
\category{F.4.1}{Mathematical Logic}{Logic and constraint programming, Mechanical theorem proving}
\category{I.2.3}{Deduction and Theorem Proving}{Inference engines, Logic programming, Resolution}

\terms
theory, verification, algorithms

\keywords
SMT-solving, static analysis, abstract interpretation, disjunctions

\section{Introduction}
Static analysis by abstract interpretation is a fully automatic program analysis method. When applied to imperative programs, it computes an inductive invariant mapping each program location (or a subset thereof) to a set of states represented symbolically~\cite{CousotCousot_JLC92}.
For instance, if we are only interested in scalar numerical program variables, such a set may be a convex polyhedron (the set of solutions of a system of linear inequalities)~\cite{CousotHalbwachs78,Halbwachs_PhD,PPL,BagnaraHZ08SCP}.

In such an analysis, information may flow forward (one computes the polyhedron after a program statement as the image by the semantics of the statement, or a super-set thereof, of the polyhedron before) or backward;
forward program analysis computes super-sets of the states reachable from the initialization of the program, backward program analysis computes super-sets of the states co-reachable from some property of interest (for instance, the violation of an assertion).
In forward analysis, control-flow joins correspond to convex hulls if using convex polyhedra (more generally, they correspond to least upper bounds in a lattice); in backward analysis, it is control-flow splits that correspond to convex hulls.

It is a known limitation of program analysis by abstract interpretation that this convex hull, or more generally, least upper bound operation, may introduce states that cannot occur in the real program: for instance, the convex hull of the intervals $[-2,-1]$ and $[1,2]$ is $[-2,2]$, strictly larger than the union of the two.
Such introduction may prevent proving desired program properties, for instance $\neq 0$. The alternative is to keep the union symbolic (e.g. compute using $[-2,-1] \cup [1,2]$) and thus compute in the \emph{disjunctive completion}
of the lattice, but the number of terms in the union may grow exponentially with the number of successive tests in the program to analyze, not to mention difficulties for designing suitable widening operators for enforcing the convergence of fixpoint iterations~\cite{PPL,BagnaraHZ08SCP,DBLP:journals/sttt/BagnaraHZ07}.
The exponential growth of the number of terms in the union may be controlled by heuristics that judiciously apply least upper bound operations, as in the \emph{trace partitioning domain} \cite{Rival_Mauborgne_TOPLAS07} implemented in the Astr\'ee analyzer~\cite{ASTREE_PLDI03,ASTREE_ESOP05}.

Assuming we are  interested in a loop-free program fragment, the above approach of keeping symbolic unions gives the same results as performing the analysis separately over every path in the fragment.
A recent method for finding disjunctive loop invariants \cite{DBLP:conf/pldi/GulwaniZ10} is based on this idea: each path inside the loop body is considered separately.
Two recent proposals use SMT-solving \cite{Kroening_Strichman_08} as a decision procedure for the satisfiability of first-order arithmetic formulas in order to enumerate only paths that are needed for the progress of the analysis \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}. They can equivalently be seen as analyses over a multigraph of transitions between some distinguished control nodes. This multigraph has an exponential number of edges, but is never explicitly represented in memory; instead, this graph is \emph{implicitly} or \emph{succinctly} represented: its edges are enumerated as needed as solutions to SMT problems.

An additional claim of the methods that distinguish paths inside the loop body \cite{DBLP:conf/pldi/GulwaniZ10,Monniaux_Gonnord_SAS11} is that they tend to generate better invariants than methods that do not, by behaving better with respect to the \emph{widening operators} \cite{CousotCousot_JLC92}
used for enforcing convergence when searching for loop invariants by Kleene iterations. A related technique, \emph{guided static analysis} \cite{DBLP:conf/sas/GopanR07}, computes successive loop invariants for increasing subsets of the transitions taken into account, until all transitions are considered; again, the claim is that this approach avoids some gross over-approximation introduced by widenings.

All these methods improve the precision of the analysis by keeping the same abstract domain (say, convex polyhedra) but changing the operations applied and their ordering. An alternative is to change the abstract domain, for instance by moving from intervals to octagons or convex polyhedra \cite{DBLP:journals/lisp/Mine06}, or the widening operator: for instance one may replace the classical widening on convex polyhedra \cite{CousotHalbwachs78,Halbwachs_PhD} by an improved version \cite{BagnaraHRZ05SCP}, or use widening ``up to'', that is, use linear inequalities syntactically present in the program source code as possible widening steps~\cite{Polka:FMSD:97}.

There are many possible combinations of the above techniques, and it is not
evident which ones perform more or less precisely or more or less efficiently on real-life examples. One needs to experiment.
Unfortunately, the published literature on the subject lacks experimental comparative assessments. One purpose of this article is therefore to propose such experimental results.

This article makes the following contributions:
\begin{enumerate}
\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph})
\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
\item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.

Our tool uses a \emph{single static assignment} (SSA) intermediate representation, which allows us to perform cheap symbolic propagations and simplifications,
described in Section~\ref{sec:analysis-algorithm}.

\item We conducted extensive experiments with this tool, on real-life programs: (i) fixing the abstract domain, varying the iteration technique (\S~\ref{sec:compare_techniques}) (ii) fixing the iteration technique, varying the abstract domain (\S\ref{sec:compare_domains}).
\end{enumerate}

\section{Bases}
\subsection{Static analysis by abstract interpretation}
\label{sec:static_analysis}
Let $X$ be the set of possible states of the program variables; for instance, if the program has 3 unbounded integer variables, then $X = \ZZ^3$. The set $\parts{X}$ of subsets of $X$, partially ordered by inclusion, is the \emph{concrete domain}. An \emph{abstract domain} is a set $\abstr{X}$ equipped with a partial order $\sqsubseteq$; for instance, it can be the domain of convex polyhedra in $\QQ^3$ ordered by geometric inclusion. In this article, all abstract domains are supposed to contain machine-representable objects, and all $\sqsubseteq$ order relations are supposed to be decidable.
The concrete and abstract domains are connected by a monotone \emph{concretization} function $\gamma: \left(\abstr{X},\sqsubseteq\right) \rightarrow (\parts{X},\subseteq)$: an element $\abstr{x} \in \abstr{X}$ represents a set $\gamma(\abstr{x})$.%
%
\footnote{%
Some presentations of abstract interpretation identify an element $\abstr{x}$ with the set of states $\gamma(\abstr{x})$ that it represents. This leads to simpler notations, but may also confuse if, as in some of our constructions, there exist several $\abstr{x}$ with identical $\gamma(\abstr{x})$, but distinct algorithmic behaviors.}

We also assume a join operator $\sqcup: \abstr{X} \times \abstr{X} \rightarrow \abstr{X}$, with infix notation; in practice, it is generally a least upper bound operation, but we only need it to satisfy $\gamma(\abstr{x}) \cup \gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \sqcup \abstr{y})$ for all $\abstr{x},\abstr{y}$.

Classically, one considers the control-flow graph of the program, with edges labeled with concrete transition relations (e.g. $x' = x+1$ for an instruction \lstinline|x = x+1;|), and attaches an abstract element to each control point.
A concrete transition relation $\tau \subseteq X \times X$ is replaced by an abstract \emph{forward abstract transformer} $\abstr{\tau}: \abstr{X} \rightarrow \abstr{X}$, such that
\begin{equation}
\forall \abstr{x} \in \abstr{X}, x,x' \in X,~
x \in \gamma(\abstr{x}) \land (x,x') \in \tau \implies
x' \in \gamma \circ \abstr{\tau}(\abstr{x})
\label{eqn:transformer_correct}
\end{equation}
It is easy to see that if to any control point $p \in P$ we attach an abstract element $\abstr{x}_p$ such that
(i) for any $p$, $\gamma(\abstr{x}_p)$ includes all initial states possible at control node $p$
(ii) for any $p,p'$, $\abstr{\tau}_{p,p'} (\abstr{x}_p) \sqsubseteq \abstr{x}_{p'}$, noting $\tau_{p,p'}$ the transition from $p$ to~$p'$, then $(\gamma(\abstr{x}_p))_{p \in P}$ form an \emph{inductive invariant}: by induction, when the control point is $p$, the program state always lies in $\gamma(\abstr{x}_p)$.

\emph{Kleene iterations} compute such an inductive invariant as the stationary limit, if it exists, of the following system: for each $p$, initialize $\abstr{x}_p$ such that $\gamma(\abstr{x}_p)$ is a superset of the initial states at point $p$; then iterate the following:
if $\abstr{\tau}_{p,p'} (\abstr{x}_p) \nsqsubseteq \abstr{x}_{p'}$, replace $\abstr{x}_{p'}$ by $\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p)$.
Such a stationary limit is bound to exist if $\abstr{X}$ has no infinite ascending chain $a_1 \sqsubsetneq a_2 \sqsubsetneq \dots$; this condition is however not met by domains such as intervals or convex polyhedra.

\emph{Widening-accelerated Kleene iterations} proceed by replacing $\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p)$ by $\abstr{x}_{p'} \widening (\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p))$ where $\widening$ is a \emph{widening operator}: for all $\abstr{x},\abstr{y}$, $\gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \widening \abstr{y})$, and there exists no sequence $\abstr{u}_1,\abstr{u}_2,\dots$ of the form $\abstr{u}_{n+1} = \abstr{u}_n \widening \abstr{v}_n$ where $\abstr{v}_n$ is another sequence.
Such iterations necessarily lead to a stationary limit $(\abstr{x}_p)_{p \in P}$, which defines an inductive invariant $(\gamma(\abstr{x}_p))_{p \in P}$. Note that this invariant is not, in general, the least one expressible in the abstract domain, and may depend on the iteration ordering (the successive choices~$p,p'$).

Once an inductive invariant $\gamma((\abstr{x}_p)_{p \in P})$ has been obtained,
one can attempt \emph{decreasing} or \emph{narrowing} iterations to reduce it.
In their simplest form, this just means running the following operation until a
fixpoint or a maximal number of iterations are reached: for any $p'$, replace
$\abstr{x}_{p'}$ by $\abstr{x}_{p'} \cap \left(\bigsqcup_{p \in P}
\abstr{\tau}_{p,p'} (\abstr{x}_p)\right)$. The result also defines an inductive
invariant. These decreasing iterations are indispensable to recover properties
from guards (tests) in the program in most iteration settings; unfortunately,
certain loops, particularly those involving identity (no-operation) transitions,
may foil them: the iterations immediately reach a fixpoint and do not decrease
further (see example in \S\ref{subsec:rate_lim}). Sections \ref{sec:guided} and
\ref{sec:path_focusing} describe techniques that work around this problem.

Widening operations have a somewhat counterintuitive behavior: the result of an analysis (postcondition) is not necessarily monotonic with respect to the precondition. For instance, classical interval analysis with widening and decreasing iterations on the following loop yields a loop invariant $x \in [0,+\infty)$ if the precondition is $x \in [0,0]$ , but $x \in [0,10]$ (strictly stronger) if the precondition is $x \in [0,10]$ (strictly weaker):
\lstinline|while (x != 10) x=x+1;|.
This means that, when using widenings, being locally more precise (for instance, by using more expressive abstract domains, or widening operators that climb more slowly \cite{BagnaraHRZ05SCP}) does not necessarily translate into higher final precision.
It is almost always possible to concoct examples where a supposedly more precise abstraction (e.g. polyhedra) yields less precise results than a less precise one (e.g. intervals), but are these examples representative of real programs?
This justifies our recourse to evaluation on realistic examples~(\S~\ref{sec:compare_techniques}).

\subsection{SMT-solving}
Boolean satisfiability (SAT) is the canonical NP-complete problem: given a propositional formula (e.g. $(a \lor \neg b) \land (\neg a \lor b \lor \neg c)$), decide whether it is satisfiable --- and, if so, output a satisfying assignment.
Despite an exponential worst-case complexity, the DPLL algorithm \cite{Kroening_Strichman_08,Handbook_SAT} solves many useful SAT problems in practice.

SAT was extended to \emph{satisfiability modulo theory} (SMT): in addition to propositional literals, SMT formulas admit atoms from a theory.
For instance, the theories of linear integer arithmetic (LIA) and linear real arithmetic (LRA) have atoms of the form $a_1 x_1 + \dots + a_n x_n \bowtie C$ where $a_1,\dots,a_n,C$ are integer constants, $x_1,\dots,x_n$ are variables (interpreted over $\ZZ$ for LIA and $\RR$ or $\QQ$ for LRA), and $\bowtie$ is a comparison operator $=,\neq,<,\leq,>,\geq$.
Satisfiability for LIA and LRA is NP-complete, yet tools based on DPLL(T) approach \cite{Kroening_Strichman_08,Handbook_SAT} solve many useful SMT problems in practice. All these tools provide a \emph{satisfying assignment} if the problem is satisfiable.

Most SMT solvers, including Z3,
%\footnote{\url{http://research.microsoft.com/en-us/um/redmond/projects/z3/}}
Yices,%
%\footnote{\url{http://yices.csl.sri.com/}}
and all those supporting the full SMTLIB2 standard \cite{BarST-SMTLIB},
offer an \emph{incremental} interface: the client program specifies the formula as an initially empty conjunction, to which additional constraints are added, and calls a ``check'' function answering whether it is satisfiable; 
it may then backtrack some of the constraints and add other ones without restarting from scratch.

\subsection{A simple, motivating example}
\label{subsec:rate_lim}
Consider the following program, adapted from \cite{Monniaux_Gonnord_SAS11}, where \lstinline|input($a$, $b$)| stands for a nondeterministic input in $[a,b]$:
\lstinputlisting{rlim_int.c}
This program implements a construct commonly found in control programs (in e.g.
automotive or avionics): a rate limiter. For the sake of simplicity, we chose it
to be fed on a nondeterministic input clamped between $[-100000,100000]$, but in a real system it would be integrated in a reactive control loop and its input connected to a complex system with unknown output range.

The expected inductive invariant is $\lstinline|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\lstinline|x_old| \in (-\infty,+\infty)$~\citep{ASTREE_ESOP05}.
Let us briefly see why.

Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
The \lstinline|x > x_old+10| test at line~6, if taken, yields $\lstinline|x_old|
\in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain
$\lstinline|x| \in (-\infty,\allowbreak 100000)$, and the same after union with the no-operation ``else'' branch. Line~7 yields $\lstinline|x| \in (-\infty,+\infty)$.

We could use ``widening up to'' or ``widening with thresholds'', propagating the ``magic values'' $\pm 100000$ associated to \lstinline|x| into \mbox{\lstinline|x_old|,} but these syntactic approaches cannot directly cope with programs for which $\lstinline|x|  \in [-100000,+100000]$ is itself obtained by analysis.
The guided static analysis of \citet{DBLP:conf/sas/GopanR07} does not perform
better, and also obtains $\lstinline|x_old| \in (-\infty,+\infty)$.

In contrast, let us distinguish all four possible execution paths through the tests at lines 6 and~7. The path through both ``else'' branches is infeasible; the program is thus equivalent to:
\lstinputlisting{rlim_int2.c}

\begin{figure}
\centering
\begin{minipage}[c]{.19\textwidth}
\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
                    semithick,font=\footnotesize]

	\node[PRstate] (n0) {$p$};
	\node[state] (n1) [below left of=n0] {};
	\node[state] (n2) [below right of=n0] {};
	\node[state] (n3) [below right of=n1] {};
	\node[state] (n4) [below left of=n3] {};
	\node[state] (n5) [below right of=n3] {};
	\node[state] (n6) [below right of=n4] {};
	\node[state] (n7) [below left of=n6] {};
	\node[state] (n8) [below right of=n6] {};
	\node[PRstate] (n9) [below right of=n7] {$q$};

  \path [transition] 
		(n0) edge              node {} (n1);
  \path [transition] 
		(n0) edge              node {} (n2);
  \path [transition] 
		(n1) edge              node {} (n3);
  \path [transition] 
		(n2) edge              node {} (n3);
  \path [transition] 
		(n3) edge              node {} (n4);
  \path [transition] 
		(n3) edge              node {} (n5);
  \path [transition] 
		(n4) edge              node {} (n6);
  \path [transition] 
		(n5) edge              node {} (n6);
  \path [transition] 
		(n6) edge              node {} (n7);
  \path [transition] 
		(n6) edge              node {} (n8);
  \path [transition] 
		(n7) edge              node {} (n9);
  \path [transition] 
		(n8) edge              node {} (n9);
\end{tikzpicture}
\end{minipage} 
$\Longrightarrow$
\begin{minipage}[c]{.19\textwidth}
\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
                    semithick,font=\footnotesize]

	\node[PRstate] (n0) {$p$};
	\node (n1) [below left of=n0] {};
	\node (n2) [below right of=n0] {};
	\node (n3) [below right of=n1] {};
	\node (n4) [below left of=n3] {};
	\node (n5) [below right of=n3] {};
	\node (n6) [below right of=n4] {};
	\node (n7) [below left of=n6] {};
	\node (n8) [below right of=n6] {};
	\node[PRstate] (n9) [below right of=n7] {$q$};

  \path [transition] 
		(n0) edge  [out=0, in=0]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=180, in=180]        node {} (n9);
  \path [transition] 
		(n0) edge  [out=205, in=155]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=230, in=130]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=255, in=105]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=-25, in=25]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=-50, in=50]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=-75, in=75]            node {} (n9);
\end{tikzpicture}
\end{minipage}
\caption{Expansion of the transition graph into a multigraph.}
\label{fig:multigraph}
\end{figure}

Classical interval analysis on this program yields $\lstinline|x_old| \in [-100000,100000]$.
We have transformed the first program into the second, manually pruning out infeasible paths; yet in general the resulting program could be exponentially larger than the first (as in Fig.~\ref{fig:multigraph}), even though not all feasible paths are needed to compute the invariant.

Following recent suggestions \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}, we avoid this space explosion by keeping the second program implicit while simulating its analysis. This means we work on an implicitly represented transition multigraph (Fig.~\ref{fig:multigraph}); it is succinctly representated by the transition graph of the first program.
Our first contribution (\S\ref{sec:guided_multigraph}) is to recast the ``guided analysis'' from \citet{DBLP:conf/sas/GopanR07} on such a succinct representation of the paths in lieu of the individual transitions.
A similar explosion occurs in disjunctive invariant generation, following \citet{DBLP:conf/pldi/GulwaniZ10}; our second contribution (\S\ref{sec:disjunctive}) applies our implicit representation to their method.

\subsection{Guided static analysis}
\label{sec:guided}
\emph{Guided static analysis} was proposed by \citet{DBLP:conf/sas/GopanR07} as an improvement over classical upward Kleene iterations with widening.
Consider the following program, taken from \cite{DBLP:conf/sas/GopanR07}:
\lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}

\begin{figure}
\begin{center}
\includegraphics[scale=0.75]{figures/gopan_reps}
\end{center}
\caption{The invariant for Prog.~\ref{lst:gopan_reps}: the piecewise linear, solid line is the strongest invariant, the grayed polyhedron is its convex hull.}
\label{fig:gopan_reps_invariant} 
\end{figure}

Classical iterations on the domain of convex polyhedra \cite{CousotHalbwachs78,BagnaraHRZ05SCP} or octagons \cite{DBLP:journals/lisp/Mine06} start with $x = 0 \land x = 0$, then continue with $x = y \land 0 \leq x \leq 1$.
The widening operator extrapolates from these two iterations and yields $x = y \land x \geq 0$.
From there, the ``else'' branch at line 5 may be taken; with further widening, $0 \leq y \leq x$ is obtained as a loop invariant, and thus the postcondition computed at line 9 is $x \geq 0 \land y = 0$.
Yet the strongest invariant is $(0 \leq x \leq 51 \land y = x) \lor (51 \leq x \leq 102 \land x+y=102)$, and its convex hull, a convex polyhedron (Fig.~\ref{fig:gopan_reps_invariant}), is
\begin{equation}
y \leq x \land y \leq 102-x \land y \geq 0.\label{eqn:triangle}
\end{equation}
Intuitively, this disappointing result is obtained because widening extrapolates from the first iterations of the loop, but the loop has two different phases ($x \leq 50$ and $x > 50$) with different behaviors, thus the extrapolation from the first phase is not valid for the second.

\citeauthor{DBLP:conf/sas/GopanR07}' idea is to analyze the first phase of the loop with a widening and narrowing sequence, and thus obtain $0 \leq x \leq 50 \land y = x$, and then analyze the second phase, finally obtaining invariant~\ref{eqn:triangle}; each phase is identified by the tests taken or not taken.

The analysis starts by identifying the tests taken and not taken during the first iteration of the loop, starting in the loop initialization. The branches not taken are pruned from the loop body, yielding:
\begin{lstlisting}[numbers=none]
  while(1) {
    if(x <= 50) y++;
    else break; /* not taken in phase 1 */
    if(y < 0) break;
    x++;
  }
\end{lstlisting}

Analyzing this loop using widening and narrowing on convex polyhedra or octagons yields the loop invariant $0 \leq x \leq 51 \land y = x$. Now, the transition at line~5 becomes feasible; and we analyze the full loop, starting iterations from $0 \leq x \leq 51 \land y = x$, and obtain invariant~(\ref{eqn:triangle}).

More generally, this analysis method considers an ascending sequence of subsets of the transitions in the loop body (left side of Fig.~\ref{fig:multigraph});
for each subset, an inductive invariant is computed for the program restricted to it.
The starting subset are the transitions reachable in one step from the loop initialization.
If for a given subset $S$ in the sequence, no transitions outside $S$ are reachable from the inductive invariant attached to $S$, then iterations stop;
otherwise, add these transitions to $S$ and iterate more.
Termination ensues from the finiteness of the control-flow graph.

\subsection{Path-focusing}
\label{sec:path_focusing}

\citet{Monniaux_Gonnord_SAS11}'s \emph{path-focusing} technique
distinguishes the different paths in the program in order to avoid loss of
precision due to merge operations. Since the number of paths may be exponential,
the technique keeps them implicit and computes them when needed using
SMT-solving.
The (accelerated) Kleene iterations (\S\ref{sec:static_analysis}) are computed over a reduced multigraph instead of the classical transition graph.

Let $P$ be the set of control points in the transition graph, 
$P_W \subseteq P$ the set of widening points
such that removing the points in $P_W$ gives an acyclic graph.
One can choose a set $P_R$ such that $P_W \subseteq P_R \subseteq P$.

The set of paths is kept implicit by an SMT formula $\rho$ expressing
the semantics of the program, assuming that the transition semantics can be
expressed within a decidable theory. For an easy construction of $\rho$, 
we also assume that the program is expressed in SSA form, meaning that each
variable is only assigned once in the transition graph. This is not a
restriction, since there exist standard algorithms that transform a program into
an SSA format.

This formula contains Boolean \emph{reachability predicates} $b_i$ for each
control points $p_i \notin P_R$, $b_i^s$ and $b_i^d$ for each $p_i \in P_R$, so
that a path 
$p_{i_1} \rightarrow p_{i_2} \rightarrow \dots \rightarrow p_{i_n}$ 
between two points $p_{i_1}, p_{i_n} \in P_R$ 
can easily be expressed as the
conjunction $b_{i_1}^s \wedge \bigwedge_{2 \leq k < n} b_{i_k} \wedge b_{i_n}^d$.
The Boolean $b_{i}^s$ is $true$ when the path starts at point $p_i$, whereas
$b_i^d$ is $true$ when the path arrives at $p_i$. In other words, we split the
points in $P_R$ into a \emph{source} point, with only outgoing transitions, and
a \emph{destination} point, with only incoming transitions, so that the
resulting graph is acyclic and there is no paths going through control
points in $P_R$.

In order to find focus paths, we solve an SMT formula which is satisfiable when
there exist a path starting at a point $p_i \in P_R$ in a state included in the
current invariant candidate $X_i$, and arriving at a point $p_j \in P_R$ in a
state outside $X_j$. In this case, we construct this path using the model and
update $X_j$. When $p_i = p_j$, meaning that the path is actually a self-loop,
we can apply a widening/narrowing sequence, or even compute the transitive
closure of the loop (or an approximation thereof, or its application to $X_i$)
using abstract acceleration~\cite{DBLP:conf/sas/GonnordH06}.

\section{Guided analysis over the paths}
\label{sec:guided_multigraph}

Guided static analysis, as proposed by \citet{DBLP:conf/sas/GopanR07}, applies to the transition
graph of the program. We now present a new technique applying this analysis on the implicit
multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
unfeasible paths.
In this section, we use the same notations as \ref{sec:path_focusing}, except
that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for
another abstract value $X_i^d$ later in the algorithm.

The combination of these two techniques aims at first discovering a precise
inductive invariant for a subset of paths between two points in $P_R$, 
by the mean of ascending and narrowing iterations. When an
inductive invariant has been found, we add new feasible paths to the subset and
compute an inductive invariant for this new subset, starting with the results
from the previous analysis.
In other words, our technique considers an ascending sequence of
subsets of the paths between two points in $P_R$.
We iterate the operations until the whole program (i.e all the
feasible paths) has been considered. The result will then be an
inductive invariant of the entire program.

\subsection{Algorithm}

Algorithm~\ref{algo:combined} performs Guided static analysis on the implicitly represented multigraph.

The current working subset of paths, noted $P$ and initially empty, is
stored using a compact representation, such as binary decision
diagrams. We also maintain two sets of control points:
\begin{itemize}
	\item $A'$ contains the points in $P_R$ that may be the starting points of new
		feasible paths.
	\item $A$ contains the points in $P_R$ on which we apply the ascending iterations.
	Each time the abstract value of a control point $p$ is updated, $p$ is
	inserted in both $A$ and $A'$.
\end{itemize}

\begin{algorithm}
	\caption{Guided static analysis on implicit multigraph}
	\label{algo:combined}
	\begin{algorithmic}[1] 
	\input{algo/algo-combined.tex}
	\end{algorithmic}
\end{algorithm}

>>>>

\documentclass{entcs} 
\usepackage{entcsmacro}
\usepackage{ifxetex}
\ifxetex
\else
\usepackage[utf8]{inputenc}
\fi

\usepackage{amsmath}
\usepackage{mathcomp,amsfonts,amssymb,MnSymbol}
\usepackage{natbib}
\usepackage[mathscr]{euscript}
\usepackage{algorithmic,algorithm}
%\usepackage{graphicx}
%\usepackage{color}

@@

\usepackage{listings}

@@

\def\newblock{\hskip .11em plus .33em minus .07em}

% The following is enclosed to allow easy detection of differences in
% ascii coding.
% Upper-case    A B C D E F G H I J K L M N O P Q R S T U V W X Y Z
% Lower-case    a b c d e f g h i j k l m n o p q r s t u v w x y z
% Digits        0 1 2 3 4 5 6 7 8 9
% Exclamation   !           Double quote "          Hash (number) #
% Dollar        $           Percent      %          Ampersand     &
% Acute accent  '           Left paren   (          Right paren   )
% Asterisk      *           Plus         +          Comma         ,
% Minus         -           Point        .          Solidus       /
% Colon         :           Semicolon    ;          Less than     <
% Equals        =3D           Greater than >          Question mark ?
% At            @           Left bracket [          Backslash     \
% Right bracket ]           Circumflex   ^          Underscore    _
% Grave accent  `           Left brace   {          Vertical bar  |
% Right brace   }           Tilde        ~

% A couple of exemplary definitions:

\newcommand{\Nat}{{\mathbb N}}
\newcommand{\Real}{{\mathbb R}}
\def\lastname{Henry}
\begin{document}
\begin{frontmatter}
  \title{PAGAI} 
  \author{Julien Henry \thanksref{imag.fr}}
  \address{VERIMAG\\ Universit\'e Joseph Fourier\\
    Grenoble, France} 
	\author{David Monniaux\thanksref{imag.fr}}
  \address{VERIMAG\\CNRS\\
    Grenoble, France} 
	\author{Matthieu Moy\thanksref{grenoble-inp.fr}}
  \address{VERIMAG\\Grenoble-INP\\
    Grenoble, France}

    \thanks[imag.fr]{Email:
    \href{First Name.Last Name@imag.fr} {\texttt{\normalshape
        First Name.Last Name@imag.fr}}} 
    \thanks[grenoble-inp.fr]{Email:
    \href{Matthieu.Moy@grenoble-inp.fr} {\texttt{\normalshape
        Matthieu.Moy@grenoble-inp.fr}}} 
\begin{abstract} 
	We present design and implementation of PAGAI, a new static analyzer 
	working over the LLVM compiler infrastructure, that computes inductive
	invariants on the numerical variables of the program.
	PAGAI implements various state-of-the-art techniques combining Abstract
	Interpretation and Decision Procedures (SMT-solving), and offers
	several abstract domains.
	We thus can compare in time and precision the various combinations of
	techniques and abstract domains. We conducted extensive experiments for
	comparing these techniques on several benchmarks and GNU programs.
\end{abstract}
\begin{keyword}
  Please list keywords from your paper here, separated by commas.
\end{keyword}
\end{frontmatter}
\section{Introduction}\label{intro}
Introduction

\section{Implementation}
